\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-12-16 archive.tex
completeness and consistency theorems are complete and consistent

@@

\system~takes as its goal to
acquire domain-specific expertise.
It gets this expertise from
solving a collection of
programming \textbf{tasks} -- typically about a hundred --
while jointly
authoring a library of code (the DSL)
and training a neural network

@@

\subsection{Refactoring code with version spaces}\label{appendixVersion}
Recall that our goal is to define an operator over version spaces, $I\beta_n$,
which calculates the set of $n$-step refactorings.
We define this operator recursively in terms of another operator, $I\beta'$, which performs a single step of refactoring:
\begin{align*}
  I\beta_n(v)& = \uplus \left\{ \underbrace{I\beta'(I\beta'(I\beta'(\cdots}_{i \text{ times}} v))) \;:\; 0\leq i \leq n \right\}
\end{align*}
where
  \begin{align*}
    I\beta'(u)& = \uplus \left\{(\lambda b)v\;:\;v\mapsto b\in S_0(u)%\text{, when }v\not=\Lambda
    \right\}\cup
  \begin{cases}
    \text{if $u$ is a primitive or index or $\varnothing $ or $\Lambda$:}&\varnothing\\
    \text{if $u = \lambda b$:}&\lambda I\beta'(b)\\
    \text{if $u = (f\;x)$:}&(I\beta'(f)\;I\beta'(x))\\
    \text{if $u = \uplus V$:}&\uplus\left\{I\beta'(u')\;|\;u'\in V \right\}
  \end{cases}  
  \end{align*}
  where we have defined $I\beta'$
  in terms of another operator, $S_k:\text{VS}\to 2^{\text{VS}\times\text{VS}}$,
  whose purpose is to construct the set of
  substitutions that are refactorings of a program in a version space.
We define $S$ as:
  \begin{align*}
  S_k(v)& = \left\{\downshift{k}_0v\mapsto \$k \right\}\cup
  \begin{cases}
    \text{if $v$ is primitive:}&\left\{\Lambda\mapsto v \right\}\\    
    \text{if $v = \$i$ and $i < k$:}&\left\{\Lambda\mapsto \$i \right\}\\
    \text{if $v = \$i$ and $i\geq k$:}&\left\{\Lambda\mapsto \$(i + 1) \right\}\\
    \text{if $v = \lambda b$:}&\left\{v'\mapsto \lambda b' \;:\; v'\mapsto b'\in S_{k + 1}(b)  \right\}\\
    \text{if $v = (f\;x)$:}&\left\{v_1\cap v_2\mapsto (f'\;x') \;:\; v_1\mapsto f'\in S_k(f),\; v_2\mapsto x'\in S_k(x) \right\}\\
    \text{if $v = \uplus V$:}&\bigcup_{v'\in V}S_n(v')\\
    \text{if $v$ is $\varnothing$:}&\varnothing\\
    \text{if $v$ is $\Lambda$:}&\left\{\Lambda\mapsto\Lambda \right\}
  \end{cases}\\
  \downshift{k}_c \$i& = \$i\text{, when $i < c$}\\
  \downshift{k}_c \$i& = \$(i - k)\text{, when $i\geq c + k$}\\
  \downshift{k}_c \$i& = \varnothing \text{, when $c\leq i <  c + k$}\\
  \downshift{k}_c \lambda b &= \lambda\downshift{k}_{c + 1}b\\
  \downshift{k}_c (f\;x)& = (\downshift{k}_cf\;\downshift{k}_cx)\\
  \downshift{k}_c \uplus V& = \uplus \left\{\downshift{k}_c v \;|\;v\in V \right\}\\
  \downshift{k}_c v& = v\text{, when }v\text{ is a primitive or }\varnothing \text{ or }\Lambda
\end{align*}
where $\shift{k}$ is the shifting operator~\cite{pierce},
which adds $k$ to all of the free variables in a $\lambda$-expression or version space,
and we have defined a new operator, $\downshift$, whose purpose is to
undo the action of $\shift$.
We have written definitions recursively,
but implement them using a dynamic program:
we hash cons each version space,
and only calculate the operators $I\beta_n$,
$I\beta'$, and $S_k$ once per each version space.
Refactoring is similarly done more quickly with dynamic programming (see Equation~\ref{FACTORING} for the recursive definition of \textsc{refactor}).

@@

From these results the consistency and completeness of $I\beta$ follows immediately by its definition.

>>>>

  \end{tikzpicture}
  \caption{}\label{threeCycles}
  \end{figure}

@@

\system~takes as its goal to
acquire domain-specific expertise,
which it learns by
solving a collection of
programming \textbf{tasks} -- typically about a hundred.
During its Wake phase,
it searches for programs solving tasks,
guided by its current DSL and recognition model.
It cycles through Wake

@@

\subsection{Refactoring code with version spaces}\label{appendixVersion}
Recall that our goal is to define an operator over version spaces, $I\beta_n$,
which calculates the set of $n$-step refactorings.
We define this operator recursively in terms of another operator, $I\beta'$, which performs a single step of refactoring:
\begin{align*}
  I\beta_n(v)& = \uplus \left\{ \underbrace{I\beta'(I\beta'(I\beta'(\cdots}_{i \text{ times}} v))) \;:\; 0\leq i \leq n \right\}
\end{align*}
where
  \begin{align*}
    I\beta'(u)& = \uplus \left\{(\lambda b)v\;:\;v\mapsto b\in S_0(u)%\text{, when }v\not=\Lambda
    \right\}\cup
  \begin{cases}
    \text{if $u$ is a primitive or index or $\varnothing $:}&\varnothing\\
    \text{if $u$ is $\Lambda$:}&\left\{\Lambda \right\}\\
    \text{if $u = \lambda b$:}&\left\{\lambda I\beta'(b) \right\}\\
    \text{if $u = (f\;x)$:}&\left\{(I\beta'(f)\;x),\;(f\;I\beta'(x)) \right\}\\
    \text{if $u = \uplus V$:}&\left\{I\beta'(u')\;|\;u'\in V \right\}
  \end{cases}  
  \end{align*}
  where we have defined $I\beta'$
  in terms of another operator, $S_k:\text{VS}\to 2^{\text{VS}\times\text{VS}}$,
  whose purpose is to construct the set of
  substitutions that are refactorings of a program in a version space.
We define $S$ as:
  \begin{align*}
  S_k(v)& = \left\{\downshift{k}_0v\mapsto \$k \right\}\cup
  \begin{cases}
    \text{if $v$ is primitive:}&\left\{\Lambda\mapsto v \right\}\\    
    \text{if $v = \$i$ and $i < k$:}&\left\{\Lambda\mapsto \$i \right\}\\
    \text{if $v = \$i$ and $i\geq k$:}&\left\{\Lambda\mapsto \$(i + 1) \right\}\\
    \text{if $v = \lambda b$:}&\left\{v'\mapsto \lambda b' \;:\; v'\mapsto b'\in S_{k + 1}(b)  \right\}\\
    \text{if $v = (f\;x)$:}&\left\{v_1\cap v_2\mapsto (f'\;x') \;:\; v_1\mapsto f'\in S_k(f),\; v_2\mapsto x'\in S_k(x) \right\}\\
    \text{if $v = \uplus V$:}&\bigcup_{v'\in V}S_n(v')\\
    \text{if $v$ is $\varnothing$:}&\varnothing\\
    \text{if $v$ is $\Lambda$:}&\left\{\Lambda\mapsto\Lambda \right\}
  \end{cases}\\
  \downshift{k}_c \$i& = \$i\text{, when $i < c$}\\
  \downshift{k}_c \$i& = \$(i - k)\text{, when $i\geq c + k$}\\
  \downshift{k}_c \$i& = \varnothing \text{, when $c\leq i <  c + k$}\\
  \downshift{k}_c \lambda b &= \lambda\downshift{k}_{c + 1}b\\
  \downshift{k}_c (f\;x)& = (\downshift{k}_cf\;\downshift{k}_cx)\\
  \downshift{k}_c \uplus V& = \uplus \left\{\downshift{k}_c v \;|\;v\in V \right\}\\
  \downshift{k}_c v& = v\text{, when }v\text{ is a primitive or }\varnothing \text{ or }\Lambda
\end{align*}
where $\shift{k}$ is the shifting operator~\cite{pierce},
which adds $k$ to all of the free variables in a $\lambda$-expression or version space,
and we have defined a new operator, $\downshift$, whose purpose is to
undo the action of $\shift$.
We have written definitions recursively,
but implement them using a dynamic program:
we hash cons each version space,
and only calculate the operators $I\beta_n$,
$I\beta'$, and $S_k$ once per each version space.
Refactoring is similarly done more quickly with dynamic programming (see Equation~\ref{FACTORING} for the recursive definition of \textsc{refactor}).

@@

From these results the consistency and completeness of $I\beta$ follows:
\begin{theorem}
  \textbf{Consistency of $I\beta'$.}\\
  If $p\in \denotation{I\beta'(u)}$ then there exists $p'\in \denotation{u}$ such that $p\reduce p'$.
\end{theorem}
\begin{proof}
  Proceed by structural induction on $u$. If $p\in \denotation{I\beta'(u)}$ then, from the definition of $I\beta'$ and $\denotation{\cdot }$, at least one of the following holds:
  \begin{itemize}
  \item Case $p = (\lambda b')v'$ where $v'\in v$, $b'\in b$, and $v\mapsto b\in S_0(u)$:
    From the definition of $\beta$-reduction we know that $p\reduce \shift{-1}[\$0\mapsto \shift{1}v']b'$.
    From the consistency of $S_n$ we know that $\shift{-1}[\$0\mapsto \shift{1}v']b'\in u$.
    Identify $p' = \shift{-1}[\$0\mapsto \shift{1}v']b'$.
  \item Case $u = \lambda b$ and $p = \lambda b'$
    where $b'\in \denotation{I\beta'(b)}$:
    By induction there exists $b''\in \denotation{b}$ such that $b'\reduce b''$.
    So $p \reduce \lambda b''$.
    But $\lambda b''\in \denotation{\lambda b} = \denotation{u}$,
    so identify $p' = \lambda b''$.
  \item Case $u = (f\; x)$ and $p = (f'\; x')$ where $f'\in \denotation{I\beta'(f)}$ and $x'\in \denotation{x}$:
    By induction there exists $f''\in \denotation{f}$
    such that $f'\reduce f''$.
    So $(f'\;x')\reduce (f''\;x')$.
    But $(f''\;x')\in \denotation{(f\;x)} = \denotation{u}$,
    so identify $p' = (f''\;x')$.
  \item  Case $u = (f\; x)$ and $p = (f'\; x')$ where $x'\in \denotation{I\beta'(x)}$ and $f'\in \denotation{f}$: Symmetric to the previous case.
  \item Case $u = \uplus U$ and $p \in \denotation{I\beta'(u')}$ where $u'\in U$:
    By induction there is a $p'\in \denotation{u'}$ satisfying $p'\reduce p$.
    But $\denotation{u'}\subseteq \denotation{u}$, so also $p'\in \denotation{u}$.
  \item Case $u$ is an index, primitive, $\varnothing$, or $\Lambda$: The theorem holds vacuously.    
    \end{itemize}  
\end{proof}
\begin{theorem}
  \textbf{Completeness of $I\beta'$.}\\
  Let $p\reduce p'$ and $p'\in \denotation{u}$.
  Then $p\in \denotation{I\beta'(u)}$.
\end{theorem}
\begin{proof}
  Structural induction on $u$.
  If $u = \uplus V$
  then there is a $v\in V$
      such that $p' \in \denotation{v}$;
      by induction on $v$ combined with the definition of $I\beta'$ we have $p\in \denotation{I\beta'(v)}\subseteq \denotation{I\beta'(u)}$,
      which is what we were to show.
      Otherwise assume that $u\not=\uplus V$.

  From the definition of $p\reduce p'$
  at least one of these cases must hold:
  \begin{itemize}
  \item Case $p = (\lambda b')v'$ and $p' = \shift{-1}[\$0\mapsto \shift{1}v']b'$:
    Using the fact that $\shift{-1}[\$0\mapsto \shift{1}v']b'\in \denotation{u}$, we can invoke the completeness of $S_n$ to construct a $(v\mapsto  b)\in S_0(u)$
    such that $v'\in \denotation{v}$
    and $b'\in \denotation{b}$.
    Combine these facts with the definition of $I\beta'$
    to get
    $p = (\lambda b')v'\in \denotation{(\lambda b)v}\subseteq I\beta'(u)$.
  \item Case $p = \lambda b$ and $p' = \lambda b'$ where
    $b\reduce b'$:
    Because $p' = \lambda b'\in \denotation{u}$
    and by assumption $u\not=\uplus V$,
    we know that $u = \lambda v$ and $b'\in \denotation{v}$. By
    induction $b\in \denotation{I\beta'(v)}$.
    Combine with the definition of $I\beta'$ to get $p = \lambda b \in \denotation{\lambda I\beta'(v)}\subseteq \denotation{I\beta'(u)}$.
  \item Case $p = (f\;x)$ and $p' = (f'\;x)$ where $f\reduce f'$:
    Because $p' = (f'\;x)\in \denotation{u}$
    and by assumption $u\not=\uplus V$
    we know that $u = (a\;b)$
    where $f'\in \denotation{a}$
    and $x\in \denotation{b}$.
    By induction on $a$
    we know
    $f\in \denotation{I\beta'(a)}$.
    Therefore
    $p = (f\;x)\in \denotation{(I\beta'(a)\; b)}\subseteq\denotation{I\beta'((a\;b))}\subseteq\denotation{I\beta'(u)}$.
  \item Case $p = (f\;x)$ and $p' = (f\;x')$ where $x\reduce x'$: Symmetric to the previous case.    
  \end{itemize}
\end{proof}
Finally we have our main result:
\begin{theorem}
  \textbf{Consistency and completeness of $I\beta_n$.}
  Let $p$ and $p'$
  be programs.
  Then $p\underbrace{\reduce q\reduce\cdots\reduce}_{\text{$\leq n$ times}} p'$  if and only
  if $p\in \denotation{I\beta_n(p')}$.
\end{theorem}
\begin{proof}
  Induction on $n$.

  If $n = 0$
  then $\denotation{I\beta_n(p')} = \left\{p \right\}$
  and $p = p'$;
  the theorem holds immediately. Assume $n > 0$.

  If $p\underbrace{\reduce q\reduce\cdots\reduce}_{\text{$\leq n$ times}} p'$
  then $q\underbrace{\reduce\cdots\reduce}_{\text{$\leq n - 1$ times}} p'$;
  induction on $n$
  gives $q\in \denotation{I\beta_{n - 1}(p')}$.
  Combined with $p\reduce q$ we can invoke the completeness of $I\beta'$
  to get $p\in \denotation{I\beta'(I\beta_{n - 1}(p'))}\subset\denotation{I\beta_n(p')}$.

  If $p\in \denotation{I\beta_n(p')}$
  then
  there exists a $i\leq n$
  such that $p\in \denotation{\underbrace{I\beta'(I\beta'(I\beta'(\cdots}_{i \text{ times}} p')))}$.
  If $i = 0$
  then $p = p'$ and $p$ reduces to $p'$ in $0\leq n$ steps.
  Otherwise $i > 0$
  and $p\in \denotation{I\beta'(\underbrace{I\beta'(I\beta'(\cdots}_{i - 1 \text{ times}} p'))}$.
  Invoking the
  consistency of $I\beta'$
  we know that $p\reduce q$
  for a program $q\in \denotation{\underbrace{I\beta'(I\beta'(\cdots}_{i - 1 \text{ times}} p'))}\subseteq\denotation{I\beta_{i - 1}(p')}$.
  By induction $q\underbrace{\reduce\cdots\reduce}_{\text{$\leq i - 1$ times}} p'$,
  which combined with $p\reduce q$
  gives  $p\underbrace{\reduce q\reduce\cdots\reduce}_{\text{$\leq i\leq n$ times}} p'$.  
\end{proof}
